Variation in use of East Asian Late Paleolithic weapons: A study of tip cross-sectional area of stemmed points from Korea

### 2023-02-03 analysis/paper/paper.qmd
prediction done

We distinguished between multiple assemblages at a single site where numerous excavations have been conducted in different locations at the site, and by different institutions under the same site name. For example, Suyanggae site, a registered Korean National Heritage site, has been excavated more than ten times since 1980 by the local university museum and later by archaeological research institutions. There are six different excavation locations that range from a few meters apart to a few kilometers. Similarly, we identified multiple assemblages in a single excavation or even a trench where archaeological deposits were separated by culturally sterile deposits, or where distinct artifact-bearing stratigraphic units could be identified by major differences in the texture, color, and composition of the sedimentary deposits. As a result, in this research we used four assemblages from the Suyanggae site. We separated one assemblage among the four by using a different site name, Hajinri, following the convention established the excavators of that location. Hajinri is the sixth excavation area at Suyanggae, which is 3.5 kilometers apart from the other areas and dated to much earlier (around 42-30 ka) than the other assemblages (around 31-15 ka). According to the excavation reports for Hajinri, stemmed points first appeared there around 42 ka with the earliest blade and blade cores [@Lee_Woo_Lee_An_Yun_Park_Otani_Kim_Kim_Han_et_al_2018]. While we include data from Hajinri here, we consider this assemblage as an unusual outlier because the stemmed points and blades from Hajinri are highly standardized and refined, which are only found at other sites much later in time. We have more confidence in finds from Bonggok, which has the second-earliest dates (41.5 ka) in our collection, as the first appearance of stemmed points. Bonggok includes blades or elongate flakes, but without any accompanying blade cores [@park2022did].

@@

Selective use of raw material is a key characteristic of the technological transition of the Korean Late Paleolithic. Prior to the Late Paleolithic, people mostly used quartzite and vein quartz for stone artefacts. Then, finer grained materials were newly added to assemblages for producing the newly introduced tools [@seong2004quartzite]. We analyzed TCSA values of `r n_raw` artifacts with raw material information to examine the interaction of raw material types and TCSA values. We categorized rare raw materials, which have less than 10 artifacts, as "Other".

@@

We used radiocarbon ages to explore the temporal patterns of the likely uses of stemmed points. After excluding assemblages that have no radiocarbon dates, we used `r n_sites_with_radiocarbon_ages` assemblages dated from 45ka to 14.8ka to explore variation in TCSA over time. We divided the artifacts from dated assemblages into three groups based on the LGM: before LGM, LGM, and after LGM, to examine the impact of this major climate event on TCSA values.

@@

We also explored spatial patterns using 16 geographical zones based geographical characteristics such as inland, coastal, major rivers, island, and major mountain ranges in addition to ecological information including temperature and precipitation [@lee2008relationship]. The sites in our sample occur in 14 of these zones: Imjin river basin (IRB), Metropolitan (MP), Central inland (CI), Kangwon coastal (KC), Choongnam coastal (CC), Southwestern inland (SWI), Upper Nagdong river basin (UNRB), WoolYoung coastal (WYC), Western Cholla (WC), Southern mountain (SM), Southeastern inland (SEI), Hyungsan Taewha coastal (HTC), Western south coastal (WSC), and Eastern south coastal (ESC). We then located individual assemblages on those two maps using longitudes and latitudes and explore TCSA values for each zone. We recognise that modern eco-regional maps may not fully resemble the Pleistocene landscape in which the points were made and used, however these are the only maps of eco-regional zones that are currently available.

@@

We predict spatial patterns... <!-- Please GP can you finish this paragraph? We need some predictions that correspond to the vegetation and geographical zones -->

@@

# all artefacts overall

@@

```{r}
#| label: fig-artifact-size-cluster
#| fig-cap: "A. Distribution of artifact weight showing three clusters. B. TCSA values for all artifacts. Artefact size classes indicated by the digits representing data point values"
#| fig-width:   4
#| fig-height:  2

@@

In order to explore spatial patterns and the relationship between the role of stemmed points and the environment, we located individual assemblages on maps that depict eco-regional zones. Our results show that stemmed points are only located in certain zones, Central Temperature Zone (CT) of vegetation map or Central inland (CI) (@fig-gis-tcsa-sites panel A & D) of eco-region map. We observed that there are statistically significant differences in the distributions of TCSA values between some zones (Vegetation zones: `r eco_lgm_test_pretty`, Ecoregion zones: `r for_lgm_test_pretty`). The Tukey's HSD results show that TCSA values from the South Temperature Zone (ST) zone are significantly different from the other two vegetation zones (@fig-gis-tcsa-sites panel C). Looking into this further, we see that there are only six artifacts from the ST zone, with only three of these having TSCA values above 175. The other three have TCSA values lower than 100, similar to the other zones (@fig-gis-tcsa-sites panel B). Given the small number of artefacts with extreme TCSA values in the ST zone, we hesitate to conclude this is an archaeologically significant pattern of different TCSA values. The Tukey HSD pairwise differences show no pairs of eco-regions with significant differences in TCSA values. Overall, our results indicate no clear patterns in the distribution of TCSA values across the vegetation and eco-regional zones.

>>>>

We distinguished between multiple assemblages at a single site where numerous excavations have been conducted in different locations at the site, and by different institutions under the same site name. For example, Suyanggae site, a registered Korean National Heritage site, has been excavated more than ten times since 1980 by the local university museum and later by archaeological research institutions. There are six different excavation locations that range from a few meters apart to a few kilometers. Similarly, we identified multiple assemblages in a single excavation or even a trench where archaeological deposits were separated by culturally sterile deposits, or where distinct artifact-bearing stratigraphic units could be identified by major differences in the texture, color, and composition of the sedimentary deposits. As a result, in this research we used four assemblages from Suyanggae. We separated one assemblage among the four by using a different site name, Hajinri, following the convention established the excavators of that location. Hajinri is the sixth excavation location at Suyanggae, which is 3.5 kilometers apart from the other areas and dated to much earlier (around 42-30 ka) than the other assemblages (around 31-15 ka). According to the excavation reports for Hajinri, stemmed points first appeared there around 42 ka with the earliest blade and blade cores [@Lee_Woo_Lee_An_Yun_Park_Otani_Kim_Kim_Han_et_al_2018]. While we include data from Hajinri here, we consider this assemblage as an unusual outlier because the stemmed points and blades from Hajinri are highly standardized and refined, which are only found at other sites much later in time. We have more confidence in finds from Bonggok, which has the second-earliest dates (41.5 ka) in our collection, as the first appearance of stemmed points. Bonggok includes blades or elongate flakes, but without any accompanying blade cores [@park2022did].

@@

Selective use of raw material is a key characteristic of the technological transition of the Korean Late Paleolithic. Prior to the Late Paleolithic, people mostly used quartzite and vein quartz for stone artifacts. Then, finer grained materials were newly added to assemblages for producing the newly introduced tools [@seong2004quartzite]. We analyzed TCSA values of `r n_raw` artifacts with raw material information to examine the interaction of raw material types and TCSA values. We categorized rare raw materials, which have less than 10 artifacts, as "Other".

@@

We used radiocarbon ages to explore the temporal patterns of the likely uses of stemmed points. After excluding assemblages that have no radiocarbon dates, we used `r n_sites_with_radiocarbon_ages` assemblages dated from 45ka to 14.8ka to explore variation in TCSA over time. We divided the artifacts from dated assemblages into three groups based on the LGM: before, during and after, to examine the impact of this major climate event on TCSA values.

@@

We also explored spatial patterns using 16 geographical zones based geographical characteristics such as inland, coastal, major rivers, island, and major mountain ranges in addition to ecological information including temperature and precipitation [@lee2008relationship]. The sites in our sample occur in 14 of these zones: Imjin river basin (IRB), Metropolitan (MP), Central inland (CI), Kangwon coastal (KC), Choongnam coastal (CC), Southwestern inland (SWI), Upper Nagdong river basin (UNRB), WoolYoung coastal (WYC), Western Cholla (WC), Southern mountain (SM), Southeastern inland (SEI), Hyungsan Taewha coastal (HTC), Western south coastal (WSC), and Eastern south coastal (ESC). We then located individual assemblages on those two maps using longitudes and latitudes and explore TCSA values for each zone. We recognize that modern eco-regional maps may not fully resemble the Pleistocene landscape in which the points were made and used, however these are the only maps of eco-regional zones that are currently available.

@@

We predict that TCSA values vary across vegetation types and geographical zones in response to heterogeneously distributed food resources. In harsh environments with lower patch productivity, similar to the case of LGM duration, we predict a further increase in the variability of TCSA values. On the other had, we predict a reduction in the range of TCSA values in affluent patches with predictable types of prey. In their study of hunter-gatherer mobility strategies, @hamilton2016ecological show that resources are abundant and predictable along coasts or lake shores, which is highly dependent on temperature and precipitation and thus hunter-gatherers often become effectively sedentary. Our prediction is that the range of TCSA values will be narrower in coastal areas such as ST and SW of vegetation zones than in inland areas such as CT.

@@

# all artifacts overall

@@

```{r}
#| label: fig-artifact-size-cluster
#| fig-cap: "A. Distribution of artifact weight showing three clusters. B. TCSA values for all artifacts. Artifact size classes indicated by the digits representing data point values"
#| fig-width:   4
#| fig-height:  2

@@

In order to explore spatial patterns and the relationship between the role of stemmed points and the environment, we located individual assemblages on maps that depict eco-regional zones. Our results show that stemmed points are only located in certain zones, Central Temperature Zone (CT) of vegetation map or Central inland (CI) (@fig-gis-tcsa-sites panel A & D) of eco-region map. We observed that there are statistically significant differences in the distributions of TCSA values between some zones (Vegetation zones: `r eco_lgm_test_pretty`, Ecoregion zones: `r for_lgm_test_pretty`). The Tukey's HSD results show that TCSA values from the South Temperature Zone (ST) zone are significantly different from the other two vegetation zones (@fig-gis-tcsa-sites panel C). Looking into this further, we see that there are only six artifacts from the ST zone, with only three of these having TSCA values above 175. The other three have TCSA values lower than 100, similar to the other zones (@fig-gis-tcsa-sites panel B). Given the small number of artifacts with extreme TCSA values in the ST zone, we hesitate to conclude this is an archaeologically significant pattern of different TCSA values. The Tukey HSD pairwise differences show no pairs of eco-regions with significant differences in TCSA values. Overall, our results indicate no clear patterns in the distribution of TCSA values across the vegetation and eco-regional zones.
